### 2024-08-07 cognitive-bias.qmd
20240807 - additional cognitive biases

As described [above](#sec-cognitiveBiasOverview), cognitive biases are systematic errors in thinking.
They lead us to be less accurate in our judgments and predictions.
Inaccuracy may not always be bad; several cognitive biases may serve to help people feel better about themselves, including [confirmation bias](#sec-cognitiveBiasesConfirmation), [overconfidence bias](#sec-cognitiveBiasesOverconfidence), optimism bias, self-serving bias, and actor-observer bias.
In fantasy football, however, cognitive biases and inaccurate judgments can lead to worse performance.
Cognitive biases are often due to the use of [heuristics](#sec-heuristics).

@@

- overconfidence bias
- confirmation bias
- in-group bias
- primacy effect bias
- recency effect bias
- hindsight bias
- loss aversion bias
- endowment bias
- bandwagon effect bias
- Dunning–Kruger effect bias

@@

@@

As a budding empiricist, you should actively seek out information that challenges or disconfirms your beliefs and work to incorporate it into your beliefs.
Do your best to go into observation, data analysis, and data interpretation with an open mind.

@@

@@

*Loss aversion bias* is the tendency to avoid losses rather than acquiring equivalent gains.
Loss aversion is exemplified when teams play conservatively so as "not to lose" instead of "to win."
In fantasy football, loss aversion may lead managers to start or hold onto underperforming high drafts for too long instead of a starting a more promising player out of fear of losing potential value from their initial investment.

### Endowment Bias {#sec-cognitiveBiasesEndowment}

*Endowment bias* is the tendency to overvalue merely because one owns it.
For instance, a manager might overvalue a player they drafted in the first round, refusing to trade them even if they could get a better-performing player in return.

@@

Applied to fantasy football, consider a situation in which you invest a lot of salary cap or a high draft pick to draft a promising player, but they repeatedly underperform.
If you continue to start the player to justify your large investment, instead of benching him in favor of a higher-performing player, you are committing the sunk cost fallacy.

>>>>

As described [above](#sec-cognitiveBiasOverview), cognitive biases are systematic errors in thinking.
They lead us to be less accurate in our judgments and predictions.
Inaccuracy may not always be bad; several cognitive biases may serve to help people feel better about themselves [@Miller2013c], including [confirmation bias](#sec-cognitiveBiasesConfirmation), [overconfidence bias](#sec-cognitiveBiasesOverconfidence), [optimism bias](#sec-cognitiveBiasesOptimism), and [self-serving bias](#sec-cognitiveBiasesSelfServing).
In fantasy football, however, cognitive biases and inaccurate judgments can lead to worse performance.
Cognitive biases are often due to the use of [heuristics](#sec-heuristics).

@@

- overconfidence bias
- optimism bias
- confirmation bias
- in-group bias
- hindsight bias
- outcome bias
- omission bias
- loss aversion bias
- primacy effect bias
- recency effect bias
- framing effect bias
- endowment effect bias
- bandwagon effect bias
- Dunning–Kruger effect bias

@@

### Optimism Bias {#sec-cognitiveBiasesOptimism}

*Optimism bias* is the tendency for people to overestimate one's likelihood of experiencing positive events and underestimate one's likelihood of experiencing negative events.
For instance a manager might overestimate the likelihood that their team will win the championship and may underestimate the likelihood that they may lose a player to injury.

@@

As a budding empiricist, you should actively seek out information that challenges or disconfirms your beliefs—particularly from trustworthy, reputable sources—and work to incorporate the information into your beliefs.
Do your best to go into observation, data analysis, and data interpretation with an open mind.

@@

### Outcome Bias {#sec-cognitiveBiasesOutcome}

*Outcome bias* is the tendency to evaluate the quality of a decision based on the eventual outcome of that decision, rather than based on the quality of the decision based on the information that was known at the time of the decision.
For instance, consider that a manager starts a lower-ranked player over a higher-ranked player because the lower-ranked player has a more favorable matchup.
Now, consider that the lower-ranked player scores fewer points than the higher-ranked player because the lower-ranked player got injured.
If the manager concludes that their decision-making process for who to start was flawed and thus not pay attention to matchups in subsequent decisions, this would reflect outcome bias.
Likewise, if the manager makes a poor-quality decision but gets lucky, and they attribute their success to a sound decision-making process, this would also reflect outcome bias.

Focus on the decision-making process when evaluating the quality of decisions.
Yes, also evaluate the outcomes of decisions, but do not give yourself too much credit for lucky decisions or too much blame for unlucky decisions (like injuries).
That said, one thing you can count on is that players get injured.
So, it is to your benefit to draft and compose your team to have some bench depth so you can handle injuries when they inevitably occur.

### Self-Serving Bias {#sec-cognitiveBiasesSelfServing}

*Self-serving bias* is the tendency to perceive oneself in an overly positive manner.
Doing so may function to bolster people's self-esteem.
One way that self-serving bias can commonly manifest is the tendency people to attribute successes to internal factors—such as one's character, ability, effort, or actions—but to attribute failures to external factors beyond one's control—such as others, unfairness, bad luck, etc.
That is, self-serving bias involves claiming more reponsibility for one's success than for one's failure [@Miller2013c].
For instance, a manager who wins the league may attribute their success to their fantasy football skill and smart decisions (rather than luck).
However, if they get last place, they may attribute their failure not to themselves but instead to factors outside of their control, such as bad luck, player injuries, and poor performance by players due to circumstances out of their control (e.g., bad weather conditions, coaching decisions).

Although the self-serving bias may help us feel better about ourselves, it can lead us to [overestimate our abilities](#sec-cognitiveBiasesOverconfidence) when things go well and to fail to learn from mistakes when things go poorly.
Remember that nearly half of the variability in fantasy football performance is estimated to be luck, and the other half likely due to skill [@Getty2018].
Thus, successes in fantasy football likely involve a good amount of luck; likewise, failures in fantasy football can also reflect mistakes and poor decision making that arise from lower skill.

### Omission Bias {#sec-cognitiveBiasesOmission}

*Omission bias* is the tendency to prefer inaction over action.
For instance, consider a manager who is on the fence to drop an underperforming player and to pick up someone else to start.
If the manager does not drop the player and the player scores few points, the manager is likely to be less critical of their decision than if they had picked up another player who then scored the same (low) number of points.

Consider another example adapted from @Miller2013c: two players scored the same number of points in a given week.
A Wide Receiver scored 1.6 points from 16 receiving yards on 1 target.
A Running Back scored 1.6 points from 56 receiving yards and two dropped fumbles.
The manager may harbor more blame for the player who performed a harmful action (losing two fumbles) than the player who received only one target (harmful inaction).

Pay attention to players' usage and opportunities, not just how many points a player scored and whether they had more fumbles or interceptions than other players.
The more times a player touches the ball, the more points they are likely to score.

*Loss aversion bias* is the tendency to avoid losses rather than acquiring equivalent gains.
That is, loss aversion (similar to punishment sensitivity, as opposed to reward sensitivity) is the tendency to weigh losses (or punishment) more heavily than gains (or rewards).
Loss aversion is different from risk aversion.
Risk aversion is the tendency to prefer outcomes with low uncertainty (compared to outcomes with greater uncertainty).
Loss aversion is exemplified when teams play conservatively so as "not to lose" instead of "to win."
In fantasy football, loss aversion may lead managers to start or hold onto underperforming high drafts for too long instead of a starting a more promising player out of fear of losing potential value from their initial investment.

@@

### Framing Effect Bias {#sec-cognitiveBiasesFraming}

*Framing effect bias* is the tendency to make decisions based on how information is framed (e.g., based on positive or negative connotations), rather than just the information itself.
For instance, consider a weekly column written by two different fantasy football expert commentators.
Expert A may write that a Running Back has scored at least 15 fantasy points in 70% of games this season (i.e., positive framing).
Expert B may write that the same Running Back has scored fewer than 15 fantasy points in 30% of games this season (i.e., negative framing).
Both statements convey the same statistical information; however, a manager might be more inclined to start the player when reading the statement from Expert A than from Expert B because of how the information was framed (i.e., presented).
People prefer to make decisions to avoid loss, consistent with [loss aversion bias](#sec-cognitiveBiasesLossAversion), so the negative framing might lead a person to be less likely to start the player.

Note how information is framed, knowing that the framing itself could lead you astray.
Thus, try to focus on the information itself rather than the framing when making decisions.

### Endowment Effect Bias {#sec-cognitiveBiasesEndowment}

*Endowment effect bias* is the tendency to overvalue merely because one owns it.
That is, a manager tends to value the same player more when they play for their team rather than an opponent's team.
When trading, the endowment effect bias might lead a manager to demand more to trade away a player than to acquire the same player [@@Miller2013c].
For instance, a manager might overvalue a player they drafted in the first round, refusing to trade them even if they could get a better-performing player in return.

A player's trade value is not fixed.
It makes sense to adjust a player's trade value based on many factors such as your team's needs, the needs of your trade partner's team, the player's remaining strength of schedule, etc.
However, you want to avoid adjusting a player's value based merely on the fact that he is on your team.
Evaluate a variety of trade analyzers to more objectively evaluate players' worth.

@@

Applied to fantasy football, consider a situation in which you invest a lot of salary cap or a high draft pick to draft a promising player, but they repeatedly underperform.
If you continue to start the player to justify your large investment, instead of benching him in favor of a higher-performing player, you are committing the sunk cost fallacy.
It is important to own up to our mistakes; when you make a mistake, the sooner you realize it, own it—for instance, by either benching or dropping the weak-performing player—and move on, the sooner you will be able to replace him with a better-performing player.
At the same time, you probably do not want to give too much weight to a player's performance in any given week, due to random chance that may lead a player to greatly under- or overperform; and whose future peformance is likely to [regress (whether positively or negatively) to their mean](#sec-fallaciesRegression).
